---
title: "Everything you've (n)ever wanted to know about penalty kicks: statistical models"
date: "2026-06-14"
format:
  html:
    code-fold: true
    fig-format: svg
    fig-dpi: 300
    theme:
      - cosmo
      - custom.scss
    highlight-style: monokai
execute:
  error: true
  message: false
  warning: false
---

After the [descriptives of part 1](/blog/penalty-kicks-descriptives/index.qmd), we turn to the statistical models. I came up with five questions that were of interest to me:

1. Where in the goal does a penalty have the best chance of going in?[^come-on]
2. Which takers are best at sending the goalkeeper the wrong way, and which keepers read them best?
3. Who are the best and worst penalty takers and penalty-stopping goalkeepers overall?
4. How good are teams at selecting their penalty takers?
5. How many penalties does a player need to take before their own history tells us more than the global average?

[^come-on]: I realize that because this is so obvious it may read like I have never watched or played football and am trying to learn about the game through numbers only like an American hedge fund that has just bought a football club. Unfortunately, I have spent more time watching (dire) football than I like to admit. I've also played quite a bit of football. I pose the question because it corresponds to a fun model and creates a beautiful visualization.

::: {.callout-note .key-findings title="At a glance"}

::: {.kf-section}
**Sending keeper wrong way**

- The keepers dive the wrong way in **56%** of penalties.
- Max Kruse is the best at sending the goalkeeper the wrong way, sending the keeper the wrong way in **25 of his 28** penalties (**89%**). The model estimates that he manages this **12.7pp** more often versus the average goalkeeper than the average player versus the average goalkeeper.
- Ante Budimir and Neymar are best at this from players selected for the World Cup 2026.
- Darren Bennet is the worst (**4/19** scored, for **-10.6pp**)
- Youri Tielemans is the worst among players active at WC 2026, sending the goalkeeper the wrong way only on **5 (29%)** of his 17 attempts (**+5.2pp**).

:::

::: {.kf-section}
**Penalty taking skill**

- Harry Kane is the best penalty taker (**97 scored out of 107 (91%)**) by considerable margin (**1.3pp** more added than the second best). The model estimates that he adds **8.2pp** above the average penalty taker against the average goalkeeper.
- Chicarito is the worst penalty taker (**8 scored out of 17 (47%)**). Because of regularisation the model believes his 'true' conversion rate is **-6.4pp** than the average penalty taker.
- Uncertainty remains large: for only **3** players the 95% credible interval excluding the average.
- Uncertainty is especially large for the worst penalty takers: teams do not wait for complete certainty as switching your taker is so cheap.
:::

::: {.kf-section}
**Penalty stopping skill**

- Tim Melia is the best goalkeeper (**26/53** not scored, **49%**). Also by a considerable margin (**1.4pp** more added). The model estimates that the average penalty taker scores **9.4pp** less of their penalties against him.
- Among the active goalkeepers, Dayne St Clair is the best (**16/35** saved) for **5.9pp** less scoring chance versus average taker. He is in goal for Canada this World Cup.
- Roman Bürki is the worst penalty stopper, conceding **44 from 47** penalty kicks (**94%**). Part of this was bad luck as the model estimates that the average penalty taker scores **4.6pp** more from their penalties versus him.
:::

::: {.kf-section}
**Placement preference**

- A taker's own placement history only outpredicts the global average (informed by foot) after about **30** penalties.
- Players seem to differ from one another mainly in how often they go down the middle, not in how often they use their weak side.
:::

::: {.kf-section}
**Team selection quality**

- Teams are good at picking penalty takers: the correlation between a player's likelihood of being selected and their scoring skill is estimated at 0.71 using data from the English top level and cup games.
:::

::: {style="margin-top: 10px;"}

:::
:::


# A simple expected penalty goals model based on shot placement and kicker's foot
We model whether an on-target penalty is scored from the kicker's foot and where the ball crosses the goal line:

$$
\begin{aligned}
\text{Goal}_i &\sim \operatorname{Bernoulli}(p_i) \\
\operatorname{logit}(p_i) &= \beta_0 + \beta_1\,\mathbb{1}[\text{foot}_i = \text{Right}] + f_{\text{foot}_i}(x_i, y_i),
\end{aligned}
$$

where $(x_i, y_i)$ is where the ball crosses the goal line and $f_{\text{Left}}$, $f_{\text{Right}}$ are thin-plate spline surfaces fit separately per foot (the `by = kick_foot` term). The index $i$ runs over on-target shots only — scored or saved — since shots that miss or strike the frame are deterministically no-goal and carry no placement information here.

```{r}
#| code-summary: "Read data"
library(tidyverse)
source("functions_features.R")
source("functions_plotting.R")

df_male <- nanoparquet::read_parquet(
  "data/penalties_all_seasons.parquet"
) |>
  convert_opta_to_meters() |>
  add_features() |>
  filter(!is_female_league)
```
```{r}
#| code-summary: "Fit model and calculate marginal effects"
# fit model
psxg_interaction <- brms::brm(
  formula = is_goal ~ kick_foot +
    s(shot_x_meters, shot_y_meters, by = kick_foot),
  # only shots on goal: post and wide are deterministically no-goal here
  data = df_male |> filter(outcome %in% c("Goal", "Saved")),
  family = bernoulli(link = "logit"),
  backend = 'cmdstanr',
  cores = 4,
  chains = 4,
  file = "models/brms_psxg_interaction"
)

# prepare marginal effects calculation
x_range <- c(post_left, post_right)
y_range <- c(0, crossbar_height)

# n_y is a resolution parameter. the heatmap is a raster, so we sample the fitted
# surface on a grid of cells and colour each one. we could sample every cm of the
# goal (~730x240 cells), but there are two reasons not to. first, it buys nothing:
# s(x, y) is a *smooth* spline with no cm-scale detail to resolve, so a fine raster
# looks identical to a coarse one (and heatmaps read as smooth long before per-pixel
# resolution). second, it is expensive in a way the cell count hides: predictions()
# evaluates each cell on *every posterior draw* and then summarises -- that is what
# produces the credible interval -- so a cm grid with ~4000 draws is hundreds of
# millions of evaluations to hold in memory. 80 rows is smooth on screen while
# keeping that posterior summary cheap; bump n_y if you ever want it crisper.
#
# keep cells roughly square: the goal is ~3x wider than tall
n_y <- 80
n_x <- round(n_y * diff(x_range) / diff(y_range))

# evaluate at cell *centres* so the raster fills the goal frame exactly and
# doesn't bleed half a cell past the posts/line into the grass
cell_centres <- function(rng, n) {
  edges <- seq(rng[1], rng[2], length.out = n + 1)
  utils::head(edges, -1) + diff(edges) / 2
}

# calculate marginal effects
psxg_grid <- marginaleffects::predictions(
  psxg_interaction,
  newdata = marginaleffects::datagrid(
    shot_x_meters = cell_centres(x_range, n_x),
    shot_y_meters = cell_centres(y_range, n_y),
    kick_foot = c("Left", "Right")
  )
) |>
  tibble::as_tibble() |>
  dplyr::select(
    kick_foot,
    shot_x_meters,
    shot_y_meters,
    prob = estimate,
    prob_lower = conf.low,
    prob_upper = conf.high
  )
```

To turn the fitted model into a picture, we lay a grid of points across the goal and read off the predicted scoring probability at each one, separately for a left- and right-footed taker. Every cell is one `marginaleffects` prediction (the posterior mean and its 95% credibility interval); the heatmap below simply colours them.
```{r}
#| code-summary: "Plot interactive heatmap"
#| column: page
#| fig-height: 4
#| fig-width: 14
#| out-width: "100%"
library(plotly)

# reversed so red marks low goal probability (where you don't want to shoot)
pal_psxg <- rev(wesanderson::wes_palette(
  "Zissou1Continuous",
  100,
  type = "continuous"
))

x_vals <- sort(unique(psxg_grid$shot_x_meters))
y_vals <- sort(unique(psxg_grid$shot_y_meters))

# reshape one foot's predictions into a z-matrix (rows = y, cols = x) plus a
# matching matrix of hover labels carrying the credible interval
foot_layer <- function(foot) {
  d <- psxg_grid |>
    dplyr::filter(kick_foot == foot) |>
    dplyr::arrange(shot_y_meters, shot_x_meters)
  to_mat <- function(v) {
    matrix(v, nrow = length(y_vals), ncol = length(x_vals), byrow = TRUE)
  }
  list(
    z = to_mat(d$prob),
    text = to_mat(sprintf(
      paste0(
        "Goal probability: <b>%.1f%%</b><br>",
        "95%% credible interval: %.1f%%–%.1f%%<br>",
        "x: %.2f m   y: %.2f m"
      ),
      d$prob * 100,
      d$prob_lower * 100,
      d$prob_upper * 100,
      d$shot_x_meters,
      d$shot_y_meters
    ))
  )
}

left <- foot_layer("Left")
right <- foot_layer("Right")
z_range <- range(psxg_grid$prob)

# colorbar ticks: pretty interior values, but always pin a tick at the data min
# and max so *both* ends of the scale are labelled (plotly's auto ticks label
# the top but not the minimum)
cb_tickvals <- {
  interior <- pretty(z_range)
  interior <- interior[interior > z_range[1] & interior < z_range[2]]
  c(z_range[1], interior, z_range[2])
}

# reversed Zissou1 as a plotly colorscale (red = low probability)
colorscale <- Map(
  function(i, col) list((i - 1) / (length(pal_psxg) - 1), col),
  seq_along(pal_psxg),
  pal_psxg
)

# goal frame, goal line and dive zones as static shapes
r <- diameter_post / 2
post_rect <- function(x_centre) {
  list(
    type = "rect",
    layer = "above",
    x0 = x_centre - r,
    x1 = x_centre + r,
    y0 = 0,
    y1 = crossbar_height + r,
    fillcolor = "black",
    line = list(width = 0)
  )
}
dive_line <- function(x) {
  list(
    type = "line",
    layer = "above",
    x0 = x,
    x1 = x,
    y0 = 0,
    y1 = crossbar_height,
    line = list(color = "gray50", width = 1, dash = "dash")
  )
}
goal_shapes <- list(
  post_rect(post_left),
  post_rect(post_right),
  list(
    type = "rect",
    layer = "above",
    x0 = post_left - r,
    x1 = post_right + r,
    y0 = crossbar_height - r,
    y1 = crossbar_height + r,
    fillcolor = "black",
    line = list(width = 0)
  ),
  list(
    type = "line",
    layer = "above",
    x0 = -(post_offset + 0.3),
    x1 = post_offset + 0.3,
    y0 = 0,
    y1 = 0,
    line = list(color = "darkgreen", width = 2)
  ),
  dive_line(dive_zone_left_offset),
  dive_line(dive_zone_right_offset)
)

heatmap_trace <- function(p, layer, visible) {
  add_trace(
    p,
    x = x_vals,
    y = y_vals,
    z = layer$z,
    text = layer$text,
    type = "heatmap",
    hoverinfo = "text",
    visible = visible,
    colorscale = colorscale,
    zmin = z_range[1],
    zmax = z_range[2],
    colorbar = list(
      title = list(text = "Goal probability", side = "top"),
      tickformat = ".0%",
      tickmode = "array",
      tickvals = cb_tickvals,
      orientation = "h",
      x = 0.5,
      xanchor = "center",
      y = 1.05,
      yanchor = "bottom",
      thickness = 10,
      len = 0.3,
      lenmode = "fraction"
    )
  )
}

# a ball that follows the cursor
# so the hover doubles as a to-scale sense of how big the ball is in the goal. it
# is appended as the next shape after the goal frame and moved on hover via JS.
ball_shape_index <- length(goal_shapes)
ball_js <- paste0(
  "function(el, x) {\n",
  "  var r = ",
  ball_radius,
  ";\n",
  "  var idx = ",
  ball_shape_index,
  ";\n",
  "  var ball = {type: 'circle', xref: 'x', yref: 'y',\n",
  "    x0: 0, x1: 0, y0: 0, y1: 0, visible: false, layer: 'above',\n",
  "    fillcolor: 'rgba(255,255,255,0.92)', line: {color: '#222222', width: 1.5}};\n",
  "  var init = {}; init['shapes[' + idx + ']'] = ball;\n",
  "  Plotly.relayout(el, init);\n",
  "  el.on('plotly_hover', function(d) {\n",
  "    var pt = d.points[0], u = {};\n",
  "    u['shapes[' + idx + '].x0'] = pt.x - r;\n",
  "    u['shapes[' + idx + '].x1'] = pt.x + r;\n",
  "    u['shapes[' + idx + '].y0'] = pt.y - r;\n",
  "    u['shapes[' + idx + '].y1'] = pt.y + r;\n",
  "    u['shapes[' + idx + '].visible'] = true;\n",
  "    Plotly.relayout(el, u);\n",
  "  });\n",
  "  el.on('plotly_unhover', function(d) {\n",
  "    var u = {}; u['shapes[' + idx + '].visible'] = false;\n",
  "    Plotly.relayout(el, u);\n",
  "  });\n",
  "}"
)

# height is pinned in pixels: with the 1:1 scaleanchor lock, plotly otherwise
# stretches the y-range to fill a tall container, leaving huge empty bands above
# and below the (wide, short) goal. width stays responsive (out-width 100%).
plot_ly(height = 400) |>
  heatmap_trace(left, visible = TRUE) |>
  heatmap_trace(right, visible = FALSE) |>
  layout(
    title = list(
      text = "Post-shot goal probability across the goal mouth",
      y = 0.97,
      yref = "container",
      yanchor = "top",
      x = 0.5
    ),
    hovermode = "closest",
    # symmetric l/r so the (centred) data area lines up under the centred title
    margin = list(t = 95, b = 35, l = 45, r = 45),
    xaxis = list(
      title = "",
      range = c(-(post_offset + 0.05), post_offset + 0.05),
      zeroline = FALSE,
      showgrid = FALSE,
      constrain = "domain"
    ),
    yaxis = list(
      title = "",
      range = c(-0.25, crossbar_height + 0.05),
      zeroline = FALSE,
      showgrid = FALSE,
      scaleanchor = "x",
      scaleratio = 1
    ),
    shapes = goal_shapes,
    updatemenus = list(list(
      type = "buttons",
      direction = "right",
      x = 0.02,
      xanchor = "left",
      y = 1.05,
      yanchor = "bottom",
      buttons = list(
        list(
          method = "restyle",
          label = "Left-footed",
          args = list("visible", list(TRUE, FALSE))
        ),
        list(
          method = "restyle",
          label = "Right-footed",
          args = list("visible", list(FALSE, TRUE))
        )
      )
    ))
  ) |>
  htmlwidgets::onRender(ball_js)
```

# What penalty takers are best at sending the goalkeeper the wrong way?
There is, however, more to penalties than shooting. For example, there's also the art of deceiving the goalkeeper on where you're going to shoot, or perhaps waiting until the goalkeeper chooses a side himself -- the only thing left to do then is to pass the ball into the part of the goal where the goalkeeper is not. 

Fortunately, we have information on what side the goalkeeper went and as such we can build a simple model of this skill by regressing whether the keeper went the wrong way on a taker and a goalkeeper random intercepts. The random intercepts are essential here as football and penalty taking is a very low frequency activity and this pools the information and thereby regularizes the estimates for each player and goalkeeper. Both the taker and keeper are modelled in the same model so that their skills are estimated simultaneously. 

In math terms this means that for penalty $i$ taken by $\text{taker}[i]$ against $\text{gk}[i]$,

$$
\begin{aligned}
\text{WrongWay}_i &\sim \operatorname{Bernoulli}(p_i) \\
\operatorname{logit}(p_i) &= \beta_0 + u_{\text{taker}[i]} + v_{\text{gk}[i]} \\
u_j &\sim \operatorname{Normal}(0, \sigma^2_{\text{taker}}), \qquad
v_k \sim \operatorname{Normal}(0, \sigma^2_{\text{gk}}).
\end{aligned}
$$

$\beta_0$ is the global log-odds the keeper goes the wrong way; the crossed random intercepts $u_j$ and $v_k$ are each taker's and each keeper's deviation from it. Their shrinkage toward $0$ is the regularisation we want: a player with a handful of penalties is pulled close to average, while the partial pooling lets a well-observed player move further out. The "added probability" we rank on is that deviation put back on the probability scale, $\operatorname{logit}^{-1}(\beta_0 + u_j) - \operatorname{logit}^{-1}(\beta_0)$.

```{r}
#| code-summary: "Helper functions for the skill models"
# --- Random-effect summaries from a brms binary GLMM ----------------------
# Helpers for summarising a single grouping factor in a logit model that has
# a global intercept and `(1 | group)` structure. They standardise the
# downstream column names (`base_prob`, `group_prob`, `added_prob`,
# `n_total`, `n_success`) so downstream plot/table code is reusable.

#' Posterior draws of a random effect's added probability.
#' Returns one row per draw per group level, with `added_prob =
#' plogis(b_Intercept + r_<group>) - plogis(b_Intercept)`.
extract_added_prob_draws <- function(model, group_col) {
  re_param_name <- paste0("r_", group_col)
  # spread_draws() uses its own NSE indexing grammar -- `r_<group>[<group>, ]`
  # declares the index column to spread -- which the dplyr .data pronoun can't
  # express, so this one argument is built with rlang. Everything downstream
  # reads columns through .data[[ ]].
  re_arg <- rlang::parse_expr(paste0(re_param_name, "[", group_col, ", ]"))

  draws <- rlang::inject(
    tidybayes::spread_draws(model, b_Intercept, !!re_arg)
  )

  draws |>
    dplyr::mutate(
      base_prob = plogis(b_Intercept),
      group_prob = plogis(b_Intercept + .data[[re_param_name]]),
      added_prob = group_prob - base_prob
    )
}

#' Median of `plogis(b_Intercept)` over the posterior — a single
#' representative baseline probability for plot annotations.
compute_baseline_prob <- function(draws) {
  draws |>
    dplyr::ungroup() |>
    dplyr::distinct(.draw, b_Intercept) |>
    dplyr::summarise(p = stats::median(plogis(b_Intercept))) |>
    dplyr::pull(p)
}

#' Per-group counts from the model frame (so levels match brms' R-name
#' format and any rows brms dropped don't poison the sums). Returns
#' `<group_col>`, `n_total`, `n_success`, with spaces in the group name
#' replaced by dots to match the random-effect names.
compute_group_counts <- function(model_data, group_col, outcome_col) {
  model_data |>
    dplyr::group_by(.data[[group_col]]) |>
    dplyr::summarise(
      n_total = dplyr::n(),
      n_success = sum(.data[[outcome_col]]),
      .groups = "drop"
    ) |>
    dplyr::mutate(dplyr::across(
      dplyr::all_of(group_col),
      \(x) stringr::str_replace_all(x, " ", "\\.")
    ))
}

#' Summary stats (median, mean and 2.5/97.5% quantiles of added_prob)
#' joined with counts, plus a rank and display label. `sort_by` selects which
#' statistic the rows are arranged by; `desc` chooses the direction (1 = best).
compute_ranks <- function(
  draws,
  group_col,
  counts_df,
  sort_by = "median_added",
  desc = TRUE
) {
  ranks <- draws |>
    dplyr::group_by(.data[[group_col]]) |>
    dplyr::summarise(
      median_added = stats::median(added_prob),
      mean_added = mean(added_prob),
      q025_added = stats::quantile(added_prob, probs = 0.025, names = FALSE),
      q975_added = stats::quantile(added_prob, probs = 0.975, names = FALSE),
      .groups = "drop"
    ) |>
    dplyr::left_join(counts_df, by = group_col)

  ranks <- if (desc) {
    dplyr::arrange(ranks, dplyr::desc(.data[[sort_by]]))
  } else {
    dplyr::arrange(ranks, .data[[sort_by]])
  }

  # rank is the row's position in the sorted order (1 = best); surface it in the
  # label so every row on the plot is tagged with its overall standing
  ranks |>
    dplyr::mutate(
      rank = dplyr::row_number(),
      label = sprintf(
        "%d. %s (n = %d/%d, prop = %.0f%%)",
        rank,
        stringr::str_replace_all(.data[[group_col]], "\\.", " "),
        n_success,
        n_total,
        100 * n_success / n_total
      )
    )
}

#' Build a small inline SVG "half-eye": a posterior density slab over a 95%
#' interval bar + median dot, all on a shared x-axis [xrange] so rows compare
#' directly. A dashed line marks 0 (average). `fill` colours the slab.
.dist_sparkline_svg <- function(values, xrange, fill, w = 130, h = 30) {
  d <- stats::density(values, from = xrange[1], to = xrange[2], n = 64)
  yn <- d$y / max(d$y) # normalise each slab to its own height
  q <- stats::quantile(values, c(0.025, 0.5, 0.975), names = FALSE)
  sx <- function(x) (x - xrange[1]) / diff(xrange) * w
  top <- 3
  slab_base <- h - 10
  iv_y <- h - 5
  sy <- function(t) slab_base - t * (slab_base - top)
  curve <- paste(sprintf("%.1f,%.1f", sx(d$x), sy(yn)), collapse = " ")
  parts <- c(
    sprintf(
      '<line x1="%.1f" y1="2" x2="%.1f" y2="%.1f" stroke="#aaa" stroke-width="0.7" stroke-dasharray="2,2"/>',
      sx(0), sx(0), h - 2
    ),
    sprintf(
      '<polygon points="%.1f,%.1f %s %.1f,%.1f" fill="%s" fill-opacity="0.9" stroke="%s" stroke-width="0.6"/>',
      sx(xrange[1]), slab_base, curve, sx(xrange[2]), slab_base, fill, fill
    ),
    sprintf(
      '<line x1="%.1f" y1="%.1f" x2="%.1f" y2="%.1f" stroke="#333" stroke-width="1.4"/>',
      sx(q[1]), iv_y, sx(q[3]), iv_y
    ),
    sprintf('<circle cx="%.1f" cy="%.1f" r="1.9" fill="#222"/>', sx(q[2]), iv_y)
  )
  sprintf(
    '<svg width="%d" height="%d" viewBox="0 0 %d %d" xmlns="http://www.w3.org/2000/svg">%s</svg>',
    w, h, w, h, paste(parts, collapse = "")
  )
}

#' Three stacked panels (shared x-axis) summarising the spread of per-player
#' added-probability effects: a histogram of takers vs. goalkeepers on top, then
#' one boxplot per group with the best and worst `n_label` of each labelled.
plot_skill_distribution <- function(
  taker_ranks,
  gk_ranks,
  x_label,
  taker_name_col = "taker_name",
  gk_name_col = "gk_name",
  taker_fill = "#78B7C5",
  gk_fill = "#E1AF00",
  n_label = 3,
  binwidth = 0.005
) {
  prep <- function(ranks, name_col, role) {
    ranks |>
      dplyr::transmute(
        role = role,
        player = stringr::str_replace_all(.data[[name_col]], "\\.", " "),
        median_added = median_added
      )
  }
  takers <- prep(taker_ranks, taker_name_col, "Takers")
  gks <- prep(gk_ranks, gk_name_col, "Goalkeepers")
  combined <- dplyr::bind_rows(takers, gks)
  combined$role <- factor(combined$role, levels = c("Takers", "Goalkeepers"))
  fills <- c(Takers = taker_fill, Goalkeepers = gk_fill)

  # shared x-range (probability units) so all three panels line up
  xlim <- range(combined$median_added)
  xlim <- xlim + c(-1, 1) * 0.05 * diff(xlim)

  pp_axis <- ggplot2::scale_x_continuous(
    labels = scales::label_number(scale = 100)
  )
  zero_line <- ggplot2::geom_vline(
    xintercept = 0, linetype = "dashed", color = "gray50"
  )

  p_hist <- combined |>
    ggplot2::ggplot(ggplot2::aes(median_added, fill = role)) +
    ggplot2::geom_histogram(
      binwidth = binwidth, position = "identity", alpha = 0.55
    ) +
    zero_line +
    pp_axis +
    ggplot2::scale_y_continuous(expand = ggplot2::expansion(mult = c(0, 0.05))) +
    ggplot2::scale_fill_manual(values = fills, name = NULL) +
    ggplot2::coord_cartesian(xlim = xlim) +
    ggplot2::labs(x = NULL, y = "Count") +
    ggplot2::theme_minimal() +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      panel.grid.major.x = ggplot2::element_blank(),
      axis.text.x = ggplot2::element_blank(),
      axis.ticks.x = ggplot2::element_blank(),
      legend.position = "top",
      legend.justification = "left"
    )

  # one boxplot per group, the best and worst `n_label` labelled; the group's
  # name rides the (coloured) y-axis title so the two panels are easy to tell
  # apart without leaning on the legend
  box_panel <- function(d, role_fill, show_x) {
    labelled <- dplyr::bind_rows(
      dplyr::slice_max(d, median_added, n = n_label),
      dplyr::slice_min(d, median_added, n = n_label)
    )
    ggplot2::ggplot(d, ggplot2::aes(x = median_added, y = 0)) +
      # a normal boxplot: box + whiskers summarise the pack and the boxplot's
      # own outliers are drawn as small neutral points; only the best and worst
      # few -- labelled on top -- get the group's distinct colour
      ggplot2::geom_boxplot(
        width = 0.5,
        fill = "gray90", color = "gray40", linewidth = 0.4,
        outlier.color = "gray70", outlier.size = 0.8, outlier.alpha = 0.7
      ) +
      ggplot2::geom_point(data = labelled, size = 1.5, color = role_fill) +
      ggrepel::geom_text_repel(
        data = labelled,
        ggplot2::aes(label = player),
        size = 3.3, direction = "both", nudge_y = 0.35,
        force = 6, force_pull = 0.3, box.padding = 0.4, point.padding = 0.2,
        segment.size = 0.3, segment.color = "gray50", segment.curvature = -0.1,
        arrow = grid::arrow(length = grid::unit(0.006, "npc"), type = "closed"),
        min.segment.length = 0.1, max.overlaps = Inf, seed = 42
      ) +
      zero_line +
      pp_axis +
      ggplot2::coord_cartesian(xlim = xlim, ylim = c(-0.4, 1.6)) +
      ggplot2::labs(x = if (show_x) x_label else NULL, y = unique(d$role)) +
      ggplot2::theme_minimal() +
      ggplot2::theme(
        panel.grid.minor = ggplot2::element_blank(),
        panel.grid.major.y = ggplot2::element_blank(),
        axis.text.y = ggplot2::element_blank(),
        axis.ticks.y = ggplot2::element_blank(),
        axis.title.y = ggplot2::element_text(color = role_fill, face = "bold"),
        axis.text.x = if (show_x) {
          ggplot2::element_text()
        } else {
          ggplot2::element_blank()
        },
        axis.ticks.x = if (show_x) {
          ggplot2::element_line()
        } else {
          ggplot2::element_blank()
        }
      )
  }

  patchwork::wrap_plots(
    p_hist,
    box_panel(takers, taker_fill, show_x = FALSE),
    box_panel(gks, gk_fill, show_x = TRUE),
    ncol = 1,
    heights = c(4, 2, 2)
  )
}

#' Interactive, sortable ranking table built from a compute_ranks() output and
#' its posterior `draws`. Each row shows an inline half-eye (density slab + 95%
#' interval) on a shared x-axis, slab-filled with a reversed Zissou1 colour
#' scaled to the player's median effect (blue = better than average, red =
#' worse; `higher_is_better = FALSE` flips this for goalkeepers). The exact
#' p2.5 / median / p97.5 numbers sit alongside. Shows `page_size` rows per page,
#' default-sorted best-first (first page = top 10); every numeric column is
#' sortable, so a reader can re-rank by the lower or upper bound.
table_re <- function(
  ranks_df,
  draws,
  name_col,
  effect_label = "Added probability vs. average",
  rate_label = "Observed %",
  count_label = "Made",
  higher_is_better = TRUE,
  page_size = 10,
  accent = "#78B7C5"
) {
  # shared x-axis covering every row's 95% interval and 0, lightly padded
  xrange <- range(c(ranks_df$q025_added, ranks_df$q975_added, 0))
  xrange <- xrange + c(-1, 1) * 0.04 * diff(xrange)

  # reversed Zissou1, scaled symmetrically about 0 and oriented so "good" is
  # always the blue end regardless of whether higher or lower is better
  pal <- rev(wesanderson::wes_palette("Zissou1", 100, type = "continuous"))
  sgn <- if (higher_is_better) 1 else -1
  lim <- max(abs(ranks_df$median_added))
  fill_for <- function(m) {
    i <- round((sgn * m / lim + 1) / 2 * (length(pal) - 1)) + 1
    pal[min(max(i, 1L), length(pal))]
  }

  # one inline half-eye SVG per group, from that group's posterior draws
  svgs <- draws |>
    dplyr::ungroup() |>
    dplyr::group_by(.data[[name_col]]) |>
    dplyr::summarise(
      dist = .dist_sparkline_svg(
        added_prob, xrange, fill_for(stats::median(added_prob))
      ),
      .groups = "drop"
    )

  tbl <- ranks_df |>
    dplyr::left_join(svgs, by = name_col) |>
    dplyr::transmute(
      rank = rank,
      player = stringr::str_replace_all(.data[[name_col]], "\\.", " "),
      kicks = n_total,
      made = n_success,
      obs_rate = n_success / n_total,
      dist = dist,
      lo = q025_added,
      mid = median_added,
      hi = q975_added
    )

  # signed percentage formatter; the cell keeps the raw numeric value so the
  # column still sorts numerically (reactable sorts on the data, not the label)
  added_col <- function(header, bold = FALSE) {
    reactable::colDef(
      name = header,
      # added_prob is a difference of two probabilities, so its unit is
      # percentage points, not percent
      cell = function(value) sprintf("%+.1f pp", 100 * value),
      align = "right",
      # wide enough that the header keeps its sort arrow on one line
      minWidth = 90,
      style = if (bold) list(fontWeight = 600) else NULL
    )
  }

  # size a numeric column to its header text plus room for the sort arrow, so
  # the header (see defaultColDef) stays on one line without making the column
  # wider than its longest label needs -- short-label tables would otherwise
  # carry so much dead width the whole table overflows into a horizontal scroll
  header_width <- function(label, floor = 56) {
    max(floor, round(nchar(label) * 7.5) + 30)
  }

  reactable::reactable(
    tbl,
    defaultSorted = list(rank = "asc"),
    defaultPageSize = page_size,
    showPageSizeOptions = TRUE,
    pageSizeOptions = c(10, 25, 50, 100),
    searchable = TRUE,
    resizable = TRUE,
    highlight = TRUE,
    compact = TRUE,
    # keep every header (and the sort arrow reactable injects into it) on one
    # line rather than letting it wrap onto a second row
    defaultColDef = reactable::colDef(
      headerStyle = list(whiteSpace = "nowrap")
    ),
    columnGroups = list(
      reactable::colGroup(
        name = paste0(effect_label, " (95% CrI)"),
        columns = c("dist", "lo", "mid", "hi")
      )
    ),
    columns = list(
      rank = reactable::colDef(
        name = "#", width = 60, align = "right",
        style = list(whiteSpace = "nowrap")
      ),
      player = reactable::colDef(name = "Player", minWidth = 160),
      kicks = reactable::colDef(
        name = "Kicks", width = header_width("Kicks"), align = "right"
      ),
      made = reactable::colDef(
        name = count_label, width = header_width(count_label), align = "right"
      ),
      obs_rate = reactable::colDef(
        name = rate_label,
        format = reactable::colFormat(percent = TRUE, digits = 0),
        width = header_width(rate_label),
        align = "right"
      ),
      dist = reactable::colDef(
        name = "Distribution",
        html = TRUE,
        sortable = FALSE,
        align = "center",
        minWidth = 150
      ),
      lo = added_col("p2.5"),
      mid = added_col("Median", bold = TRUE),
      hi = added_col("p97.5")
    ),
    theme = reactable::reactableTheme(
      highlightColor = "#eef6f8",
      headerStyle = list(borderBottom = paste0("2px solid ", accent))
    )
  )
}
```

```{r}
#| code-summary: "Fit model and calculate marginal effects"
m_wrong_way <- brms::brm(
  formula = gk_wrong_way ~ 1 + (1 | taker_name) + (1 | gk_name),
  data = df_male,
  family = brms::bernoulli(link = "logit"),
  backend = "cmdstanr",
  chains = 4,
  cores = 4,
  seed = 2026,
  file = "models/brms_wrong_way_skill"
)

draws_taker_ww <- extract_added_prob_draws(m_wrong_way, "taker_name")
draws_gk_ww <- extract_added_prob_draws(m_wrong_way, "gk_name")

baseline_prob_ww <- compute_baseline_prob(draws_taker_ww)

taker_counts_ww <- compute_group_counts(
  m_wrong_way$data,
  "taker_name",
  "gk_wrong_way"
)
gk_counts_ww <- compute_group_counts(
  m_wrong_way$data,
  "gk_name",
  "gk_wrong_way"
)

# higher added_prob = more often deceives the GK = better deceiver
taker_ranks_ww <- compute_ranks(
  draws_taker_ww,
  "taker_name",
  taker_counts_ww,
  desc = TRUE
)
# lower added_prob = goes the wrong way less often = better at reading
gk_ranks_ww <- compute_ranks(
  draws_gk_ww,
  "gk_name",
  gk_counts_ww,
  desc = FALSE
)
```

```{r}
#| code-summary: "Plot results takers"
#| column: page
table_re(
  taker_ranks_ww,
  draws_taker_ww,
  "taker_name",
  effect_label = "Added wrong-way probability vs. average",
  rate_label = "Wrong-way %",
  count_label = "Wrong-way"
)
```

At the top of this list we see players with a small stutter like Lewandowski and Bruno Fernandes. The very best, however, Max Kruse, manages to [wait and misdirect](https://www.youtube.com/watch?v=aGHc7lApO_g) the keeper without a noticeable stutter even!

Messi, for all his body feints in open play, is not great at misdirecting the goalkeeper in penalties -- he ranks `r taker_ranks_ww |>  mutate(rank = row_number()) |> filter(str_detect(taker_name, "Lionel.Messi")) |>  pull(rank)` out of `r nrow(taker_ranks_ww)`, sending goalkeepers the wrong way only `r taker_ranks_ww |> filter(str_detect(taker_name, "Lionel.Messi")) |> mutate(p = scales::percent(n_success / n_total, accuracy = 1)) |> pull(p)` of the time!

# What goalkeepers are best at reading takers (or just wait (too) long)?
```{r}
#| code-summary: "Plot results goalkeepers"
#| column: page
table_re(
  gk_ranks_ww,
  draws_gk_ww,
  "gk_name",
  effect_label = "Added wrong-way probability vs. average",
  rate_label = "Wrong-way %",
  count_label = "Wrong-way",
  higher_is_better = FALSE
)
```

You can also use this table to find goalkeepers that often dive the correct way, but still fail to save the penalties. They either wait too long or are just bad at stopping the ball (or are just unlucky).

```{r}
#| code-summary: "Plot distribution of wrong-way skill"
plot_skill_distribution(
  taker_ranks_ww,
  gk_ranks_ww,
  x_label = "Added wrong-way probability vs. average (pp)"
)
```

Note how much wider this skill distribution is for penalty takers than for goalkeepers.

# Quantifying penalty taking and stopping skill from the top down
Now, modeling all skills from the bottom up is hard as you then need to identify each of the skills that go into penalty taking. Even if you were able to write down all skills that go into this, you would also need the data on these skills. However, we do not have data on the speed at which the ball travels and I expect many of the skills to be latent. Therefore, I opt to model the skill from the top down. The model is very similar as before:

$$
\begin{aligned}
\text{Goal}_i &\sim \operatorname{Bernoulli}(p_i) \\
\operatorname{logit}(p_i) &= \beta_0 + u_{\text{taker}[i]} + v_{\text{gk}[i]} \\
u_j &\sim \operatorname{Normal}(0, \sigma^2_{\text{taker}}), \qquad
v_k \sim \operatorname{Normal}(0, \sigma^2_{\text{gk}}).
\end{aligned}
$$

It is the same crossed random-intercept structure as the wrong-way model, the outcome now being whether the penalty was scored: $u_j$ is taker $j$'s scoring skill and $v_k$ is keeper $k$'s effect on it (a better stopper pushes the probability down), each relative to the global log-odds $\beta_0$.

```{r}
#| code-summary: "Fit model"
m <- brms::brm(
  formula = is_goal ~ (1 | taker_name) + (1 | gk_name),
  data = df_male,
  family = brms::bernoulli(link = "logit"),
  backend = "cmdstanr",
  chains = 4,
  cores = 4,
  seed = 2026,
  file = "models/brms_penalty_skill"
)
```

It's a deliberately simple model: we do not take possible differences in context between penalty histories from players into account. So no information on derbies, knock-out, shootouts, chance to go ahead etc. -- just the taker, the goalkeeper, and whether the penalty was scored. Note the outcome is whether the taker *scored*, not whether the keeper *saved*. Part of a keeper's skill may be unsettling the taker into missing the target altogether, and modelling "scored" lets the keeper take credit for that too.

## Now, among penalty takers, who are the best? Who are the worst? 
```{r}
#| code-summary: "Calculate marginal effects"
library(tidybayes)

draws_taker <- extract_added_prob_draws(m, "taker_name")
draws_gk <- extract_added_prob_draws(m, "gk_name")

baseline_prob <- compute_baseline_prob(draws_taker)

taker_counts <- compute_group_counts(m$data, "taker_name", "is_goal")
gk_counts <- compute_group_counts(m$data, "gk_name", "is_goal")

# higher added_prob = more likely to score = better taker
taker_ranks <- compute_ranks(
  draws_taker,
  "taker_name",
  taker_counts,
  desc = TRUE
)
# lower added_prob = fewer goals conceded = better gk
gk_ranks <- compute_ranks(draws_gk, "gk_name", gk_counts, desc = FALSE)
```

```{r}
#| code-summary: "Plot results penalty takers"
#| column: page
table_re(
  taker_ranks,
  draws_taker,
  "taker_name",
  effect_label = "Added scoring probability vs. average",
  rate_label = "Scored %",
  count_label = "Scored"
)
```

Harry Kane stands out as the best penalty taker in the dataset by a considerable margin. Remarkably, [he missed](https://youtu.be/Q-zmXSb0WKU?t=103) one of the highest-profile penalties in his career when he blazed his penalty over the top thereby failing to equalize for England versus France in the 2022 World Cup quarter-final.

We can also see exactly why Thomas Tuchel brought a striker playing in the Saudi League into the squad: Ivan Toney ranks as an exceptionally strong penalty taker.

Interestingly, for only `r sum(taker_ranks$q025_added > 0)` players, the credibility interval excludes 0. This starkly demonstrates just how many penalties a model requires to establish absolute certainty about a player's skill level.[^caveat_skill] It is almost hard to believe that we can confidently label so few players as "above average," which may suggest the need for a more informative prior.

[^caveat_skill]: They have data from training though and I believe there are other qualities that are informative on penalty skill, e.g. ball striking. It's probably not a coincidence that Harry Kane is on top as he is a historically great ball striker. On the other hand, [other historical good finishers](https://blogarchive.statsbomb.com/articles/soccer/quantifying-finishing-skill/), like Griezman, Messi, Luis Suárez, and Son Heung-Min are only average or even worse than average penalty takers.

Finally, it is also worth noting that the uncertainty in estimated skill is bigger for the worst takers because -- of course -- their environment picks up on their ability and they won't be the designed penalty taker for their team for long -- preventing the accumulation of many penalties across their career. Switching penalty takers is cheap. 

```{r}
#| code-summary: "Plot distribution of penalty-taking and -stopping skill"
plot_skill_distribution(
  taker_ranks,
  gk_ranks,
  x_label = "Added scoring probability vs. average (pp)"
)
```

Takers (blue) and goalkeepers (gold) on the same scale. Both groups bunch tightly around the average once the sparsely-observed are shrunk toward it; only a handful -- the proven specialists and the likely liabilities -- pull away.

## Who are the best penalty stopping goalkeepers, who are the worst? 
```{r}
#| code-summary: "Plot results goalkeepers"
#| column: page
table_re(
  gk_ranks,
  draws_gk,
  "gk_name",
  effect_label = "Added scoring probability vs. average",
  rate_label = "Conceded %",
  count_label = "Conceded",
  higher_is_better = FALSE
)
```

Among active goalkeepers, Bart Verbruggen ranks `r gk_ranks |> filter(str_detect(gk_name, "Verbruggen")) |> pull(rank)` of `r nrow(gk_ranks)` (`r gk_ranks |> filter(str_detect(gk_name, "Verbruggen")) |> transmute(s = paste0(scales::number(median_added, scale = 100, accuracy = 0.1, style_positive = "plus"), " pp, 95% CrI ", scales::number(q025_added, scale = 100, accuracy = 0.1, style_positive = "plus"), " to ", scales::number(q975_added, scale = 100, accuracy = 0.1, style_positive = "plus"))) |> pull(s)`) and Robin Roefs `r gk_ranks |> filter(str_detect(gk_name, "Roefs")) |> pull(rank)` of `r nrow(gk_ranks)` (`r gk_ranks |> filter(str_detect(gk_name, "Roefs")) |> transmute(s = paste0(scales::number(median_added, scale = 100, accuracy = 0.1, style_positive = "plus"), " pp, 95% CrI ", scales::number(q025_added, scale = 100, accuracy = 0.1, style_positive = "plus"), " to ", scales::number(q975_added, scale = 100, accuracy = 0.1, style_positive = "plus"))) |> pull(s)`).

Perhaps we should do a goalkeeper substitution again with the 3rd Dutch GK looking very, very promising while the starting goalkeeper looks to be quite bad.[^dutch_gk]

[^dutch_gk]: Luck would have it that the very few penalties from Bart Verbruggen's professional career that are not in the dataset (Conference League, not covered by my data source), he saved: on February 23, 2023 [he saved 3 penalties](https://www.uefa.com/uefaconferenceleague/video/027e-1763da99f9d2-d77cd1537c85-1000--uecl-laufenn-potw-spieler-st8-bart-verbruggen-anderlecht-v-/) in a shootout for Anderlecht. I may try to obtain a more complete dataset in the lead up to the knockout phase of the tournament.

# How good are teams at selecting penalty takers? Evidence from English football
As mentioned before, there's information in the team's decision. We can model these as a latent belief about a teammate's penalty-taking skill. Most penalties are taken by the supposedly better takers, so there is a strong selection effect: we only ever observe a scoring outcome for the player who was *chosen* to take the kick. If that choice tracks skill, the takers we observe are a non-random, better-than-average slice of all footballers, and a model fit only to them mistakes "the average taker" for "the average player".

To correct for this we fit a joint (Heckman-style) model with two linked equations. For each of the players on the pitch at penalty $i$ (indexed $\text{pl}[i]$, facing keeper $\text{gk}[i]$):

$$
\begin{aligned}
\textbf{Selection:}\quad \text{Taker}_i &\sim \operatorname{Bernoulli}(\pi_i), \\
\operatorname{logit}(\pi_i) &= \gamma_{\text{slot}[i]} + \delta\,\mathbb{1}[\text{GK}_i] + a_{\text{pl}[i]} \\[4pt]
\textbf{Outcome:}\quad \text{Goal}_i \mid \text{Taker}_i{=}1 &\sim \operatorname{Bernoulli}(p_i), \\
\operatorname{logit}(p_i) &= \beta_0 + u_{\text{pl}[i]} + v_{\text{gk}[i]} \\[4pt]
\textbf{Player effects:}\quad \begin{pmatrix} a_j \\ u_j \end{pmatrix} &\sim \operatorname{Normal}(\mathbf 0,\ \Sigma),
\qquad
\Sigma = \begin{pmatrix} \sigma_a^2 & \rho\,\sigma_a\sigma_u \\ \rho\,\sigma_a\sigma_u & \sigma_u^2 \end{pmatrix}, \\[4pt]
\textbf{Keeper effect:}\quad v_k &\sim \operatorname{Normal}(0,\sigma_v^2).
\end{aligned}
$$

The _selection_ equation asks, among the eleven players on the pitch (the risk set), who actually steps up: $\gamma_{\text{slot}}$ gives a separate base rate for open-play penalties and for shootouts (where several players take a kick), $\delta$ lets goalkeepers almost never be the taker[^gk_modeling], and $a_j$ is player $j$'s latent propensity to be chosen. The _outcome_ equation is observed only for the taker (`subset(is_taker)` in brms): $\beta_0$ is the global log-odds of scoring, $u_j$ is taker $j$'s scoring skill and $v_k$ is keeper $k$'s suppression of it.

[^gk_modeling]: The result of this is that in a model fitted on the full dataset, goalkeepers like Rogério Ceni that regularly took penalties were believed to be extremely good even if their penalty records were merely good.

The two per-player effects couple the equations by drawing $a_j$ and $u_j$ from a shared bivariate normal with correlation $\rho$. $\rho$ measures whether the players teams *choose* are also the players who *score*: a positive $\rho$ says selection tracks skill. It also allows us to estimate the skill for players we never see take a penalty; their scoring skill $u_j$ is inferred from their selection propensity $a_j$ through $\rho$.  $\beta_0$ now corresponds to the average footballer rather than the average penalty taker.

Ideally, we would fit a more complex model that also takes into account _who_ the other players on the field were: if both Kane and Tony are on the field and Kane takes it, we get some information on who is the better penalty taker between them. Similarly, in shootouts there is information on who is selected among the first 5 takers. I think such a model would amount to a discrete choice model. However, it would require me to write custom `STAN` code so for now I revert to the simpler selection model that I can fit in `brms`.

```{r}
#| code-summary: "Transform data"
lineups_penalties <- get_penalty_lineups()

model_df_raw <- lineups_penalties |>
  # join the penalty event context to all 11 players on the field during penalty. 
  # inner_join so lineup rows with no matching penalty event -- i.e. the female-league
  # events filtered out of df_male -- are dropped explicitly here rather than
  # surviving as NA rows that get implicitly swept out by the slot filters below.
  # relationship = "many-to-one": many lineup players map to a single penalty
  # event (df_male is unique on match_id x event_id); dplyr errors if violated.
  inner_join(
    df_male |>
      select(
        match_id,
        event_id,
        gk_id,
        competition,
        competition_type_detailed,
        is_goal_actual = is_goal,
        is_shootout,
        shootout_team_kick_seq_nr,
        shootout_seq_total
      ),
    by = c("match_id", "event_id"),
    relationship = "many-to-one"
  )

# Collapse shootouts so each contributes one row per starter (instead of
# n_kicks * 11). For takers we keep the row of their first kick; non-takers'
# kick-level fields (is_goal_actual, gk_id, event_id) come from an arbitrary
# kick row but go unused; the outcome equation subsets on is_taker.
shootout_rows <- model_df_raw |>
  filter(is_shootout) |>
  group_by(match_id, player_id, player_name) |>
  arrange(desc(is_taker), shootout_team_kick_seq_nr, .by_group = TRUE) |>
  slice(1) |>
  ungroup()

model_df_all_comps <- bind_rows(
  model_df_raw |> filter(!is_shootout),
  shootout_rows
) |>
  # 4. Build the 'slot' variable and the masked 'is_goal'
  mutate(
    # With shootouts collapsed to (shootout x player), slot is now a 2-level
    # opportunity type: in_game vs shootout.
    slot = case_when(
      !is_shootout ~ "in_game",
      is_shootout ~ "shootout",
      TRUE ~ "unknown"
    ),
    is_gk = position == 'GK',
    # Convert slot to a factor for brms
    slot = factor(slot, levels = c("in_game", "shootout")),
    # The outcome is ONLY observed for the taker. Everyone else gets NA.
    # We must ensure is_goal is integer 1/0 for the Bernoulli family in brms.
    is_goal = if_else(is_taker, as.integer(is_goal_actual), NA_integer_),
    is_taker = as.integer(is_taker)
  ) |>
  # Drop events where we couldn't determine the slot properly
  filter(slot != "unknown")

# English top-flight competitions: the only ones in this dataset with decent cup
# (hence shootout) coverage, which is what supplies players who actually take
# penalties and so identifies the outcome equation. Reused in the prose below.
england_competitions <- c("ENG-Premier League", "ENG-FA Cup", "ENG-League Cup")

model_df_england <- model_df_all_comps |>
  filter(competition %in% england_competitions)

# Share of players on the pitch who never take a penalty in a given opportunity
# set. Never-takers carry no outcome information, so a set dominated by them is
# poor for identification.
share_never_taker <- function(df) {
  df |>
    group_by(player_id) |>
    summarise(ever_taker = any(is_taker == 1), .groups = "drop") |>
    summarise(p = mean(!ever_taker)) |>
    pull(p)
}
pct_never_england <- share_never_taker(model_df_england)
pct_never_top5_cup <- share_never_taker(
  model_df_all_comps |>
    filter(competition_type_detailed %in% c("top 5 league", "cup"))
)

```

Moreover, we limit ourselves to the English top level competitions from our dataset because the cup coverage in this dataset from the other leagues is limited. And I think this is important for model identification: `r scales::percent(pct_never_england, accuracy = 0.1)` has never taken a penalty in the English top level competitions (Premier League, FA Cup and League Cup), while this is as high as `r scales::percent(pct_never_top5_cup, accuracy = 0.1)` if we filter down to top 5 leagues and cups (because cup games where shootouts can happen are mostly not covered).

```{r}
#| code-summary: "Fit model"
library(brms)

# Equation 1: Selection (Who takes it from players on field?)
eq_selection <- bf(
  is_taker ~ 0 + slot + is_gk + (1 | rho | player_id),
  family = bernoulli()
)

# Equation 2: Outcome (Do they score?)
eq_outcome <- bf(
  is_goal | subset(is_taker) ~ 1 + (1 | rho | player_id) + (1 | gk_id),
  family = bernoulli()
)

# Combine, define priors, and fit
# Note: set_rescor(FALSE) is required for categorical/bernoulli multivariate models
joint_formula <- eq_selection + eq_outcome + set_rescor(FALSE)

priors <- c(
  # 1. Fixed Effects (Coefficients for 'slot')
  # Using a regularizing Normal or Student-t instead of the 2008 Cauchy
  prior(normal(0, 5), class = "b", resp = "istaker"),

  # 2. Intercept for the Outcome Equation
  # Slightly wider, akin to the 2008 recommendation for intercepts
  prior(normal(0, 5), class = "Intercept", resp = "isgoal"),

  # 3. Variance Components (Random Effects)
  # Using the modern Exponential instead of the 2006 half-Cauchy
  prior(exponential(1), class = "sd", resp = "istaker"),
  prior(exponential(1), class = "sd", resp = "isgoal"),

  # 4. Correlation Matrix
  prior(lkj(2), class = "cor")
)

penalty_selection_model <- brm(
  formula = joint_formula,
  data = model_df_england,
  prior = priors,
  file = "models/brms_penalty_selection_correlation_england",
  seed = 2026,
  chains = 4,
  cores = 4,
  refresh = 5,
  iter = 2000,
  backend = "cmdstanr"
)
```

```{r}
#| code-summary: wrangle top takers
library(tidybayes)

draws_taker <- spread_draws(
  penalty_selection_model,
  b_isgoal_Intercept,
  r_player_id__isgoal[player_id, term],
  ndraws = 1000
) |>
  filter(term == "Intercept") |>
  mutate(
    base_prob = plogis(b_isgoal_Intercept),
    taker_prob = plogis(b_isgoal_Intercept + r_player_id__isgoal),
    added_prob = taker_prob - base_prob
  )

taker_ranks <- draws_taker |>
  group_by(player_id) |>
  summarise(
    median_abs = median(taker_prob),
    mean_abs = mean(taker_prob),
    q025_abs = quantile(taker_prob, probs = 0.025, names = FALSE),
    q975_abs = quantile(taker_prob, probs = 0.975, names = FALSE),
    median_added = median(added_prob),
    mean_added = mean(added_prob),
    q025_added = quantile(added_prob, probs = 0.025, names = FALSE),
    q975_added = quantile(added_prob, probs = 0.975, names = FALSE),
    .groups = "drop"
  ) |>
  arrange(desc(median_added)) |>
  mutate(rank = row_number())

player_summary <- model_df_england |>
  group_by(player_id) |>
  summarise(
    n_opportunities = n(),
    n_taken = sum(is_taker),
    n_scored = sum(is_goal, na.rm = TRUE),
    .groups = "drop"
  )

summary_ext <- taker_ranks |>
  full_join(lineups_penalties |> distinct(player_id, player_name)) |>
  full_join(player_summary)

rho_post <- brms::posterior_summary(
  penalty_selection_model,
  variable = "cor_player_id__istaker_Intercept__isgoal_Intercept"
)
# baseline = average-footballer conversion (the outcome intercept on prob scale).
baseline_conv <- median(draws_taker$base_prob)
top_taker_name <- summary_ext |>
  arrange(desc(median_abs)) |>
  slice(1) |>
  pull(player_name) |>
  str_replace_all("\\.", " ")

worst_taker <- summary_ext |>
  filter(!is.na(median_abs)) |>
  arrange(median_abs) |>
  slice(1)

toney <- summary_ext |> filter(str_detect(player_name, "Toney")) |> slice(1)
```

In this filtered dataset we see that `r top_taker_name` is still the top penalty taker. However, the bottom part is much more interesting now, as the worst penalty takers are now believed to be those players who had many opportunities but were not selected by themselves or their team in shootouts or during on field penalties. And instead of getting the average conversion rate among penalty takers, we now estimate the average conversion rate among footballers. This is slightly lower, at about `r scales::percent(baseline_conv, accuracy = 1)`, with the worst players expected to convert only around `r scales::percent(worst_taker$median_abs, accuracy = 1)`. Already good takers who were also trusted a lot, like Ivan Toney (took `r toney$n_taken` of `r toney$n_opportunities` opportunities in this dataset), are estimated to be even a bit better now.

The exact skill estimate for the bottom takers are probably up for debate: a player who never takes a penalty has no scoring record, so the model can only infer their conversion from how unlikely they are to be picked, translated into skill through the correlation $\rho$. With no data to anchor them, it is the prior rather than the evidence that decides the exact number. The numbers for the worst takers are therefore best read as an educated extrapolation.

What's less up for debate I think, is the correlation parameter between a player's likelihood of taking a penalty and their ability to score one. This comes out at `r scales::number(rho_post[, "Estimate"], accuracy = 0.01)` (95% CrI `r scales::number(rho_post[, "Q2.5"], accuracy = 0.01)` to `r scales::number(rho_post[, "Q97.5"], accuracy = 0.01)`). The players who are most likely to step up (the regular penalty takers) seem indeed the best at converting them. In other words, the teams are pretty good at selecting their penalty takers!

# How many penalties do we need before there is more information in a penalty taker's dataset than global behaviour?
Every penalty goes to one of three zones: the taker's dominant side, their non-dominant side, or the centre. Because these probabilities must sum to 100%, any player's tendency can be visualized as a single point on [a three-way simplex](https://en.wikipedia.org/wiki/Ternary_plot). 

To estimate these tendencies, we can use a Dirichlet-multinomial model, which naturally handles this three-way split by breaking the problem into two layers:

$$
\begin{aligned}
\mathbf{p}_i &\sim \operatorname{Dirichlet}(\phi\,\boldsymbol{\mu}) \\
\mathbf{y}_i &\sim \operatorname{Multinomial}(n_i,\ \mathbf{p}_i)
\end{aligned}
$$

For taker $i$, $\mathbf{y}_i = (y_{i,\text{dom}}, y_{i,\text{non}}, y_{i,\text{cen}})$ are their zone counts over $n_i$ penalties and $\boldsymbol{\mu}$ is the global mean pattern (the split: the three $\mu_z$ sum to 1). The concentration $\phi$ functions like a prior sample size: it is the number of 'pseudo-kicks' worth of evidence the global pattern $\boldsymbol{\mu}$ carries, so a player's own record only starts to dominate once they have taken more than $\phi$ penalties.
```{r}
#| code-summary: Transform data and fit model
df_player_preference <- df_male |>
  select(1:5, taker_name, kick_foot, shot_zone, shot_zone_dominance) |>
  count(taker_name, shot_zone_dominance) %>%
  pivot_wider(
    names_from = shot_zone_dominance,
    values_from = n,
    values_fill = list(n = 0) # Fill NAs with 0 for players who haven't shot in a zone
  ) %>%
  mutate(total_penalty_kicks = Dominant + Non_dominant + Centre)

fit <- brms::brm(
  formula = cbind(Dominant, Non_dominant, Centre) |
    brms::trials(total_penalty_kicks) ~ 1,
  family = brms::dirichlet_multinomial(),
  data = df_player_preference,
  backend = 'cmdstanr',
  chains = 4,
  cores = 4,
  seed = 2026,
  file = "models/brms_zone_preference"
)

summary(fit)

# the Dirichlet concentration acts like a prior sample size: the global pattern is
# worth roughly this many kicks, so a player's own history only outweighs it
# once they have taken more than that
phi_zone <- unname(brms::posterior_summary(fit, variable = "phi")[, "Estimate"])
```

To see exactly why $\phi$ represents this tipping point, we can look at the posterior estimate for a player's tendency, which forms a weighted average: $\hat{\mathbf{p}}_i = \left(\frac{\phi}{\phi + n_i}\right) \boldsymbol{\mu} + \left(\frac{n_i}{\phi + n_i}\right) \frac{\mathbf{y}_i}{n_i}$. The global baseline is weighted by $\frac{\phi}{\phi + n_i}$ and the player's empirical data is weighted by $\frac{n_i}{\phi + n_i}$. A player's individual history becomes more informative than the global dataset exactly when their personal weight strictly exceeds the global weight ($\frac{n_i}{\phi + n_i} > \frac{\phi}{\phi + n_i}$). Because the denominators are identical, they cancel out and give us the threshold: $n_i > \phi$.

Because penalty placement preferences are so alike across players, a player's first few kicks don't tell us much about their own tendencies. The fitted concentration says the global average carries the weight of about `r round(phi_zone)` kicks: it isn't until a player has taken more than `r round(phi_zone)` penalties that their own history predicts their next shot better than simply going with the global placement pattern.

A few things are worth noting on this calculation. For example, you could make the argument that this high $\phi$ value is to be expected because this data consists only of _penalty takers_ and mostly of takers who have taken lots of penalties (i.e. specialists) and these know they need to be maximally unpredictable to maximize their chances in the long run and therefore vary a lot. Among inexperienced players, this may be less the case and $\phi$ may be lower. A counterpoint to this is that in the [previous blog post](/blog/penalty-kicks-descriptives/index.qmd) we observed that there is not that great of a difference in distribution across left, centre, right between different match situations, experience, and player position.[^difference] So maybe we indeed do need that many penalties. This would suggest the study of the opponent is in part an overfitting exercise.[^concede]

[^concede]: I concede that I'm not intimately familiar enough with the preparation process to make definite statements on this. Moreover, there could be different strategies for maximizing your chance in _this_ shootout and over _all_ shootouts. Looking more into it, the preparation suggested by [this article](https://www.nytimes.com/athletic/5625975/2024/07/10/netherlands-penalty-cheat-sheet-england-pickford/) seems quite sensible although they probably still overfit in light of my findings and do not take into account the impact of knowing that they know.

[^difference]: The difference between these proportions and the plot from the [previous post](/blog/penalty-kicks-descriptives/index.qmd) is that that model included all shots, including those that missed the goal, while this model is limited to those shots on goal. 

We can use these estimated $\phi$ and $\boldsymbol{\mu}$ to set up an interactive calculator that quantitatively updates our beliefs on where a player is going to shoot.

```{r}
#| echo: false
# pass the conjugate parameters (global pattern mu and concentration phi) to OJS
mu_zone <- brms::posterior_epred(
  fit, newdata = data.frame(total_penalty_kicks = 1)
)[, 1, ] |> colMeans()
ojs_define(
  phi_cal = phi_zone,
  mu_dom = unname(mu_zone["Dominant"]),
  mu_non = unname(mu_zone["Non_dominant"]),
  mu_cen = unname(mu_zone["Centre"])
)

# also pass thinned posterior draws so the calculator can propagate the
# uncertainty in phi and mu (Dominant is the softmax reference category)
draws <- as.data.frame(fit,
  variable = c("b_muNondominant_Intercept", "b_muCentre_Intercept", "phi"))
draws <- draws[round(seq(1, nrow(draws), length.out = 400)), ]
e_non <- exp(draws$b_muNondominant_Intercept)
e_cen <- exp(draws$b_muCentre_Intercept)
den <- 1 + e_non + e_cen
ojs_define(
  phi_draws = draws$phi,
  mu_dom_draws = 1 / den,
  mu_non_draws = e_non / den,
  mu_cen_draws = e_cen / den
)
```

::: {.penalty-calc}
[Where would this taker's next penalty go?]{.calc-title}

::: {.calc-grid}
::: {.calc-inputs}
```{ojs}
//| echo: false
// one viewof returning {dom, non, cen}: full-width sliders, live value beside
// each label and no number box (hence no spinner arrows)
viewof kicks = {
  const mk = (label, value) => {
    const range = html`<input type="range" min="0" max="100" step="1" value="${value}">`;
    const val = html`<span class="kick-val">${value}</span>`;
    const row = html`<div class="kick-row"><label>${label} ${val}</label>${range}</div>`;
    range.oninput = () => { val.textContent = range.value; };
    return {range, row};
  };
  const d = mk("Dominant", 3), n = mk("Non-dominant", 3), c = mk("Centre", 3);
  const form = html`<form class="kick-sliders">${d.row}${n.row}${c.row}</form>`;
  Object.defineProperty(form, "value", {
    get: () => ({dom: +d.range.value, non: +n.range.value, cen: +c.range.value})
  });
  return form;
}
```
:::

::: {.calc-plot}
```{ojs}
//| echo: false
ternary = {
  // conjugate update, recomputed locally so this cell renders only the plot
  const num = (x) => +(Array.isArray(x) ? x[0] : x);
  const phi = num(phi_cal);
  const mu = {Dominant: num(mu_dom), "Non-dominant": num(mu_non), Centre: num(mu_cen)};
  const keys = ["Dominant", "Non-dominant", "Centre"];
  const y = {Dominant: kicks.dom, "Non-dominant": kicks.non, Centre: kicks.cen};
  const n = kicks.dom + kicks.non + kicks.cen;
  const emp = {};
  for (const k of keys) emp[k] = n > 0 ? y[k] / n : mu[k];
  const calc = {n, emp};

  // --- propagate posterior uncertainty: the taker's latent preference is a
  // mixture (over posterior draws of phi, mu) of conjugate Dirichlets, which
  // has closed-form mean Mbar and covariance Sigma (law of total covariance) ---
  const S = phi_draws.length;
  const Mbar = [0, 0, 0];
  const ms = new Array(S);
  const Cw = [[0, 0, 0], [0, 0, 0], [0, 0, 0]];
  for (let s = 0; s < S; s++) {
    const a = [
      phi_draws[s] * mu_dom_draws[s] + y.Dominant,
      phi_draws[s] * mu_non_draws[s] + y["Non-dominant"],
      phi_draws[s] * mu_cen_draws[s] + y.Centre
    ];
    const a0 = a[0] + a[1] + a[2];
    const m = [a[0] / a0, a[1] / a0, a[2] / a0];
    ms[s] = m;
    const f = 1 / (a0 + 1);
    for (let i = 0; i < 3; i++) {
      Mbar[i] += m[i] / S;
      for (let j = 0; j < 3; j++)
        Cw[i][j] += ((i === j ? m[i] * (1 - m[i]) : -m[i] * m[j]) * f) / S;
    }
  }
  const Sigma = [[0, 0, 0], [0, 0, 0], [0, 0, 0]];
  for (let i = 0; i < 3; i++)
    for (let j = 0; j < 3; j++) {
      let bt = 0;
      for (let s = 0; s < S; s++) bt += (ms[s][i] - Mbar[i]) * (ms[s][j] - Mbar[j]);
      Sigma[i][j] = Cw[i][j] + bt / S;
    }
  const postMean = {Dominant: Mbar[0], "Non-dominant": Mbar[1], Centre: Mbar[2]};

  const W = 380, H = 345, padX = 52, padTop = 30, padBot = 38;
  const corners = {
    Dominant:       [0, 0],
    "Non-dominant": [1, 0],
    Centre:         [0.5, Math.sqrt(3) / 2]
  };
  const sx = d3.scaleLinear([0, 1], [padX, W - padX]);
  const sy = d3.scaleLinear([0, Math.sqrt(3) / 2], [H - padBot, padTop]);
  const bary = (p, i) =>
    p.Dominant * corners.Dominant[i] +
    p["Non-dominant"] * corners["Non-dominant"][i] +
    p.Centre * corners.Centre[i];
  const px = (p) => [sx(bary(p, 0)), sy(bary(p, 1))];

  const svg = d3.create("svg")
      .attr("viewBox", [0, 0, W, H])
      .attr("width", W).attr("height", H)
      .attr("font-family", "Atkinson Hyperlegible, sans-serif")
      .attr("style", "max-width:100%;height:auto;overflow:visible");

  // faint interior gridlines at 25 / 50 / 75 %
  for (const t of [0.25, 0.5, 0.75]) {
    const segs = [
      [{Dominant: t, "Non-dominant": 1 - t, Centre: 0}, {Dominant: t, "Non-dominant": 0, Centre: 1 - t}],
      [{Dominant: 1 - t, "Non-dominant": t, Centre: 0}, {Dominant: 0, "Non-dominant": t, Centre: 1 - t}],
      [{Dominant: 1 - t, "Non-dominant": 0, Centre: t}, {Dominant: 0, "Non-dominant": 1 - t, Centre: t}]
    ];
    for (const [a, b] of segs) {
      const A = px(a), B = px(b);
      svg.append("line")
         .attr("x1", A[0]).attr("y1", A[1]).attr("x2", B[0]).attr("y2", B[1])
         .attr("stroke", "#e7e1d4").attr("stroke-width", 1);
    }
  }

  // triangle frame
  const tri = [corners.Dominant, corners["Non-dominant"], corners.Centre]
    .map(c => [sx(c[0]), sy(c[1])].join(","));
  svg.append("polygon")
     .attr("points", tri.join(" "))
     .attr("fill", "none").attr("stroke", "#1D323E").attr("stroke-width", 1.5);

  // vertex labels -- middle-anchored and inside the viewBox so they never clip
  for (const [txt, c, dy] of [
    ["Dominant", corners.Dominant, 20],
    ["Non-dominant", corners["Non-dominant"], 20],
    ["Centre", corners.Centre, -11]
  ]) {
    svg.append("text")
       .attr("x", sx(c[0])).attr("y", sy(c[1]) + dy)
       .attr("text-anchor", "middle").attr("font-size", 12.5)
       .attr("font-weight", 700).attr("fill", "#1D323E")
       .text(txt);
  }

  // 95% credible region for the latent preference: project the propagated
  // 3-simplex covariance to pixels (A * Sigma * A^T) and draw the ellipse
  {
    const cpx = [px({Dominant: 1, "Non-dominant": 0, Centre: 0}),
                 px({Dominant: 0, "Non-dominant": 1, Centre: 0}),
                 px({Dominant: 0, "Non-dominant": 0, Centre: 1})];
    const Ar = [[cpx[0][0], cpx[1][0], cpx[2][0]], [cpx[0][1], cpx[1][1], cpx[2][1]]];
    const quad = (r, c) => {
      let v = 0;
      for (let a = 0; a < 3; a++) for (let b = 0; b < 3; b++) v += Ar[r][a] * Sigma[a][b] * Ar[c][b];
      return v;
    };
    const cxx = quad(0, 0), cxy = quad(0, 1), cyy = quad(1, 1);
    const dsc = Math.sqrt(Math.max(0, (cxx - cyy) ** 2 / 4 + cxy * cxy));
    const l1 = (cxx + cyy) / 2 + dsc, l2 = Math.max(0, (cxx + cyy) / 2 - dsc);
    const ang = Math.abs(cxy) > 1e-9 ? Math.atan2(l1 - cxx, cxy) : (cxx >= cyy ? 0 : Math.PI / 2);
    const Mp = px(postMean), kk = 5.991; // chi-square 0.95, 2 dof
    svg.append("ellipse")
       .attr("cx", Mp[0]).attr("cy", Mp[1])
       .attr("rx", Math.sqrt(l1 * kk)).attr("ry", Math.sqrt(l2 * kk))
       .attr("transform", `rotate(${ang * 180 / Math.PI}, ${Mp[0]}, ${Mp[1]})`)
       .attr("fill", "rgba(120, 183, 197, 0.18)").attr("stroke", "#78B7C5")
       .attr("stroke-width", 1).attr("stroke-opacity", 0.55);
  }

  // shrinkage segment: global -> raw record
  if (calc.n > 0) {
    const A = px(mu), B = px(calc.emp);
    svg.append("line")
       .attr("x1", A[0]).attr("y1", A[1]).attr("x2", B[0]).attr("y2", B[1])
       .attr("stroke", "#9aa7ab").attr("stroke-width", 1).attr("stroke-dasharray", "3,3");
    svg.append("circle").attr("cx", B[0]).attr("cy", B[1]).attr("r", 4)
       .attr("fill", "#9aa7ab").attr("opacity", 0.85);
  }

  // global split (hollow)
  const G = px(mu);
  svg.append("circle").attr("cx", G[0]).attr("cy", G[1]).attr("r", 5)
     .attr("fill", "#faf8f3").attr("stroke", "#1D323E").attr("stroke-width", 1.5);

  // shrunken estimate (headline)
  const P = px(postMean);
  svg.append("circle").attr("cx", P[0]).attr("cy", P[1]).attr("r", 7)
     .attr("fill", "#78B7C5").attr("stroke", "#1D323E").attr("stroke-width", 1.2);

  return svg.node();
}
```
:::

::: {.calc-out}
```{ojs}
//| echo: false
readout = {
  // conjugate update, recomputed locally so this cell renders only the table
  const num = (x) => +(Array.isArray(x) ? x[0] : x);
  const phi = num(phi_cal);
  const mu = {Dominant: num(mu_dom), "Non-dominant": num(mu_non), Centre: num(mu_cen)};
  const fmt = d3.format(".0%");
  const keys = ["Dominant", "Non-dominant", "Centre"];
  const y = {Dominant: kicks.dom, "Non-dominant": kicks.non, Centre: kicks.cen};
  const n = kicks.dom + kicks.non + kicks.cen;
  const w = n / (n + phi);
  const emp = {};
  for (const z of keys) emp[z] = n > 0 ? y[z] / n : mu[z];

  // propagate posterior uncertainty (mixture of conjugate Dirichlets over the
  // posterior draws): closed-form posterior mean and 95% credible interval
  const S = phi_draws.length;
  const idx = {Dominant: 0, "Non-dominant": 1, Centre: 2};
  const Mbar = [0, 0, 0], within = [0, 0, 0], mm = new Array(S);
  for (let s = 0; s < S; s++) {
    const a = [
      phi_draws[s] * mu_dom_draws[s] + y.Dominant,
      phi_draws[s] * mu_non_draws[s] + y["Non-dominant"],
      phi_draws[s] * mu_cen_draws[s] + y.Centre
    ];
    const a0 = a[0] + a[1] + a[2], m = [a[0] / a0, a[1] / a0, a[2] / a0];
    mm[s] = m;
    for (let i = 0; i < 3; i++) {
      Mbar[i] += m[i] / S;
      within[i] += (m[i] * (1 - m[i]) / (a0 + 1)) / S;
    }
  }
  const post = {}, ci = {};
  for (const z of keys) {
    const i = idx[z];
    let bt = 0;
    for (let s = 0; s < S; s++) bt += (mm[s][i] - Mbar[i]) ** 2;
    const sd = Math.sqrt(within[i] + bt / S);
    post[z] = Mbar[i];
    ci[z] = [Math.max(0, Mbar[i] - 1.96 * sd), Math.min(1, Mbar[i] + 1.96 * sd)];
  }
  const calc = {keys, n, w, emp, post, ci};

  const k = calc.keys;
  const cell = (obj, z) => obj === null ? "—" : fmt(obj[z]);
  const row = (name, obj, cls = "") =>
    html`<tr class="${cls}"><td>${name}</td>${k.map(z => html`<td>${cell(obj, z)}</td>`)}</tr>`;
  const postCell = (z) => html`<td>${fmt(calc.post[z])}<span class="ci">${fmt(calc.ci[z][0])}–${fmt(calc.ci[z][1])}</span></td>`;
  return html`<div class="calc-readout">
    <p>With <b>${calc.n}</b> kick${calc.n === 1 ? "" : "s"} on record, the estimate puts
       <b>${fmt(calc.w)}</b> of the weight on this taker's own history and
       <b>${fmt(1 - calc.w)}</b> on the global split.</p>
    <table>
      <thead><tr><th></th>${k.map(z => html`<th>${z}</th>`)}</tr></thead>
      <tbody>
        ${row("Global split", mu)}
        ${row("This taker's record", calc.n > 0 ? calc.emp : null)}
        <tr class="post-row"><td>Shrunken estimate <span class="ci">95% CrI</span></td>${k.map(z => postCell(z))}</tr>
      </tbody>
    </table>
    <p class="calc-note">Dot = posterior mean; the shaded ellipse and the intervals give the 95% credible region, propagating the uncertainty in φ, μ and the taker's own record.</p>
  </div>`;
}
```
:::
:::
:::

We can take our Dirichlet model a step further and let each taker have their own placement preference instead of sharing the single global one, by adding a per-taker random intercept:

$$
\begin{aligned}
\mathbf{y}_i &\sim \operatorname{Dirichlet\text{-}multinomial}(n_i,\ \phi\,\boldsymbol{\mu}_i) \\
\mu_{i,z} &\propto \exp(\beta_z + u_{i,z}), \qquad u_{i,z} \sim \operatorname{Normal}(0, \sigma_z^2)
\end{aligned}
$$

Now each taker $i$ has their own pattern $\boldsymbol{\mu}_i$, the $u_{i,z}$ pulling it away from the global $\boldsymbol{\mu}$ (Dominant is the reference, so $\beta_{\text{dom}} = u_{i,\text{dom}} = 0$). The per-zone $\sigma_z$ say how much takers vary in each zone.
```{r}
fit_re <- brms::brm(
  formula = cbind(Dominant, Non_dominant, Centre) |
    brms::trials(total_penalty_kicks) ~ 1 + (1 | taker_name),
  family = brms::dirichlet_multinomial(),
  data = df_player_preference,
  backend = 'cmdstanr',
  chains = 4,
  cores = 4,
  seed = 2026,
  file = "models/brms_zone_preference_re"
)
```

We see that each player gets their own latent preference simplex shrunk toward the global one, and the estimated random-intercept SDs come out very uneven across zones -- much larger for the centre than for the non-dominant side -- so players differ from one another mainly in how willing they are to go down the middle, not in how often they use their weak side. 

The model also comes with an asterisk as every player only contributes a single aggregated row of counts, so a per-player random intercept and the Dirichlet concentration $\phi$ end up explaining the same between-player spread and are only weakly separable (hence the borderline Rhat and ESS). I hope it can still be suggestive, though.

Let's look at the most idiosyncratic takers (and, given enough kicks, the most predictable ones):

```{r}
#| code-summary: "Most idiosyncratic takers"
#| column: page
# brms::ranef() gives a [taker, summary-stat, coefficient] array; the
# coefficients are the per-category intercept deviations (Dominant is the
# reference category). We rank takers by how far that deviation vector sits
# from 0 -- i.e. from the global average preference -- on the latent scale.
re_est <- brms::ranef(fit_re)$taker_name[, "Estimate", ]

# brms stores RE levels with spaces swapped for dots while df_player_preference
# keeps the originals, so match on a key normalised to single spaces
norm_key <- function(x) {
  stringr::str_squish(stringr::str_replace_all(x, "[. ]+", " "))
}
idiosyncrasy <- tibble::tibble(
  taker_key = norm_key(rownames(re_est)),
  idiosyncrasy = sqrt(rowSums(re_est^2))
)

# global zone split, pooled over every penalty -- the reference shown in
# brackets and used to shade each cell by how far the player sits from it
global_split <- with(df_player_preference, c(
  Dominant = sum(Dominant),
  Non_dominant = sum(Non_dominant),
  Centre = sum(Centre)
))
global_split <- global_split / sum(global_split)

idiosyncratic_takers <- df_player_preference |>
  dplyr::mutate(taker_key = norm_key(taker_name)) |>
  dplyr::left_join(idiosyncrasy, by = "taker_key") |>
  dplyr::transmute(
    Player = taker_name,
    Kicks = total_penalty_kicks,
    Dominant = Dominant / total_penalty_kicks,
    Non_dominant = Non_dominant / total_penalty_kicks,
    Centre = Centre / total_penalty_kicks,
    Idiosyncrasy = idiosyncrasy
  )

# each cell shows the player's share with the global share in brackets and an
# accent fill proportional to |player - global|; a fixed 25pp reference keeps the
# shading comparable now that every taker is listed (incl. one-kick players whose
# raw split would otherwise blow out a data-driven scale)
dev_cap <- 0.25
zone_col <- function(global, label) {
  reactable::colDef(
    name = label,
    align = "right",
    minWidth = 110,
    cell = function(value) {
      sprintf("%.0f%% (%.0f%%)", 100 * value, 100 * global)
    },
    style = function(value) {
      list(background = scales::alpha(
        "#78B7C5", min(1, abs(value - global) / dev_cap)
      ))
    }
  )
}

reactable::reactable(
  idiosyncratic_takers,
  defaultSorted = list(Idiosyncrasy = "desc"),
  defaultPageSize = 10,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 25, 50, 100),
  compact = TRUE,
  highlight = TRUE,
  searchable = TRUE,
  columns = list(
    Player = reactable::colDef(minWidth = 160),
    Kicks = reactable::colDef(width = 80, align = "right"),
    Dominant = zone_col(global_split[["Dominant"]], "Dominant"),
    Non_dominant = zone_col(global_split[["Non_dominant"]], "Non-dominant"),
    Centre = zone_col(global_split[["Centre"]], "Centre"),
    Idiosyncrasy = reactable::colDef(
      format = reactable::colFormat(digits = 2),
      align = "right",
      minWidth = 120
    )
  ),
  theme = reactable::reactableTheme(
    highlightColor = "#eef6f8",
    headerStyle = list(
      borderBottom = "2px solid #78B7C5",
      whiteSpace = "nowrap"
    )
  )
)
```


